---
title: "Week 10: Temporal data"
date: today
date-format: "DD/MM/YY"
format: pdf
execute: 
  warning: false
  message: false
---

# Child mortality in Sri Lanka

In this lab you will be fitting a couple of different models to the data about child mortality in Sri Lanka, which was used in the lecture. Here's the data and the plot from the lecture:

```{r}
library(tidyverse)
library(here)
library(rstan)
library(tidybayes)

lka <- read_csv(here("Labs/Lab8/lka.csv"))
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```

# Fitting a linear model 

Let's firstly fit a linear model in time to these data. Here's the code to do this:

```{r}
observed_years <- lka$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)

stan_data <- list(y = lka$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  mid_year = mean(years), se = lka$se)

mod <- stan(data = stan_data,
             file = here("Labs/Lab8/lka_linear_me.stan"))

```

Extract the results:

```{r}
res1 <- mod %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year = years[t])
```



Plot the results:

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res1, aes(year, .value)) + 
  geom_ribbon(data = res1, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "Linear fit shown in black")
```

## Question 1

Project the linear model above out to 2023 by adding a `generated quantities` block in Stan (do the projections based on the expected value $\mu$). Plot the resulting projections on a graph similar to that above. 


```{r}
stan_data <- list(y = lka$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  mid_year = mean(years), se = lka$se, P = 9)

mod2 <- stan(data = stan_data,
             file = here("Labs/Lab8/Lab10_1.stan"))

```



```{r}
res2 <- mod2 %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year=years[t])

res_p2 <- mod2 %>% 
  gather_draws(mu_p[p]) %>% 
  median_qi() %>% 
  mutate(year=years[nyears]+ p)
```


```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res2, aes(year, .value)) + 
  geom_ribbon(data = res2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res_p2, aes(year, .value)) + 
  geom_ribbon(data = res_p2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, col= "red")+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "Linear fit shown in black")
```

# Random walks


## Question 2

Code up and estimate a first order random walk model to fit to the Sri Lankan data, taking into account measurement error, and project out to 2023. 


```{r}
stan_data <- list(y = lka$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  mid_year = mean(years), se = lka$se, P = 9)

mod3 <- stan(data = stan_data,
             file = here("Labs/Lab8/Lab10_2.stan"))
```


```{r}
res3 <- mod3 %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year=years[t])

res_p3 <- mod3 %>% 
  gather_draws(mu_p[p]) %>% 
  median_qi() %>% 
  mutate(year=years[nyears]+ p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res3, aes(year, .value)) + 
  geom_ribbon(data = res3, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res_p3, aes(year, .value)) + 
  geom_ribbon(data = res_p3, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, col= "red")+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "first order shown in black, projections in red")
```

## Question 3

Now alter your model above to estimate and project a second-order random walk model (RW2). 

```{r}
stan_data <- list(y = lka$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  mid_year = mean(years), se = lka$se, P = 9)

mod4 <- stan(data = stan_data,
             file = here("Labs/Lab8/Lab10_3.stan"))
```

```{r}
res4 <- mod4 %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year=years[t])

res_p4 <- mod4 %>% 
  gather_draws(mu_p[p]) %>% 
  median_qi() %>% 
  mutate(year=years[nyears]+ p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res4, aes(year, .value)) + 
  geom_ribbon(data = res4, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res_p4, aes(year, .value)) + 
  geom_ribbon(data = res_p4, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, col= "red")+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "Second order shown in black, projections in red")
```

## Question 4

Run the first order and second order random walk models, including projections out to 2023. Compare these estimates with the linear fit by plotting everything on the same graph. 

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  geom_line(data = res2, aes(year, .value), color= "red") + 
  geom_ribbon(data = res2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill ="red")+
  geom_line(data = res_p2, aes(year, .value), color= "red") + 
  geom_ribbon(data = res_p2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill= "red")+
  
  geom_line(data = res3, aes(year, .value), color = "blue") + 
  geom_ribbon(data = res3, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill ="blue")+
  geom_line(data = res_p3, aes(year, .value), color = "blue") + 
  geom_ribbon(data = res_p3, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill ="blue")+
  
  geom_line(data = res4, aes(year, .value), color = "green") + 
  geom_ribbon(data = res4, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "green")+
  geom_line(data = res_p4, aes(year, .value), color = "green") + 
  geom_ribbon(data = res_p4, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill= "green")+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "second-order RW(green), first-order RW(blue), linear fit(red)")

```



## Question 5

Rerun the RW2 model excluding the VR data. Briefly comment on the differences between the two data situations. 

```{r}
# exclude VR
excluded_VR <- lka %>% filter(source != 'VR')
observed_years <- excluded_VR$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)

```

```{r}
stan_data <- list(y = excluded_VR$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  mid_year = round(mean(years)), se = excluded_VR$se, P = 18)

mod5 <- stan(data = stan_data,
             file = here("Labs/Lab8/Lab10_3.stan"))
```

```{r}
res5 <- mod5 %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year=years[t])

res_p5 <- mod5 %>% 
  gather_draws(mu_p[p]) %>% 
  median_qi() %>% 
  mutate(year=years[nyears]+ p)
```

```{r}
ggplot(excluded_VR, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res5, aes(year, .value)) + 
  geom_ribbon(data = res5, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res_p5, aes(year, .value)) + 
  geom_ribbon(data = res_p5, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, col= "red")+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "RW2 shown in black, projections in red")
```
```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res4, aes(year, .value)) + 
  geom_ribbon(data = res4, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res_p4, aes(year, .value)) + 
  geom_ribbon(data = res_p4, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, col= "red")+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "RW2 shown in black, projections in red")
```
Without data source VR, we don't have observations in year 2016. This means the prediction is not so good compared to when we have source VR available to us. 

## Question 6

Briefly comment on which model you think is most appropriate, or an alternative model that would be more appropriate in this context. 

Looking at the plot in Q4, second-order random walk may be the best model. Since it utilized more information, has similar prediction as first-order random walk model, but with lower standard error. 